---
title: "California Public Schools Lookup Table"
author: "Ryan Estrellado"
date: "5/19/2017"
output: html_document
---

## Purpose 

The purpose of this R Notebook is to document the process used to clean the 
California Department of Education's public database of public schools. This 
can be a very useful dataset for data scientists, analysts, or school staff, 
but the format of the dataset may be overwhelming to those who don't typically 
work with a dataset this size or with this many columns. In particular, the 
dataset uses the longer fourteen digit school ID code, which can make it 
difficult for users to merge with a dataset that uses the commonly seen seven 
digit school ID code. I hope that once this dataset is cleaned, it will be a 
little easier to use for work projects, including:  

- Making interactive `leaflet` maps using the longitude and latitude files 
- Merging with other datasets, such as SEIS  
- Visualizing itinerant school assignments  
  
## Reading in the Data 

The dataset we'll be using can be found on the [California Department of 
Education's website](http://www.cde.ca.gov/ds/si/ds/pubschls.asp). A description of all the variables can also be found 
[on the CDE website](http://www.cde.ca.gov/ds/si/ds/fspubschls.asp). This data 
is updated daily, so we'll read the file in directly from the URL. 

First we'll start with reading in the data. At the time of writing, CDE makes 
the data available in a text format and as an excel file format. For this 
project we'll use the text format. 

```{r message = F}
# Load the required packages

library(tidyverse)
library(knitr)
library(stringr)
```

```{r message = F, warning = F}
original <- read_tsv("ftp://ftp.cde.ca.gov/demo/schlname/pubschls.txt")
```

You can see here that the original feel is big, both in rows and columns. 

```{r comment = NULL}
dim(original)
```
  
## Cleaning the Data

We'll clean the data up by doing the following: 

- Filter the `StatusType` column so that we only have `Active` schools  
- Remove code fields if there is a more helpful description field  
- Remove other columns school staff probably won't need when merging with this 
file  

```{r}
cde_schls <- original %>% 
  filter(StatusType == "Active") %>% 
  select(-c(AdmFName1:AdmEmail3, FundingType, Magnet, StatusType, Street:State, DOC, CharterNum, SOC, EdOpsCode, EILCode))
```

Users of this data file will typically need the seven digit CDS code and not the 
longer fourteen digit code. To make it easier to merge using the seven digit 
code, we'll create a new column called `Short_CDS`. 

```{r}
cde_schls <- cde_schls %>% 
  mutate(Short_CDS = str_sub(CDSCode, 8, 20)) %>% 
  select(CDSCode, Short_CDS, everything())
```

```{r comment = NULL}
# Preview the cleaned dataset

cde_schls
```
  
## Checking for Missing Values 

It's helpful to know if there are missing values in any of the more useful 
fields. The following are the percentage of `NA`s in some of the fields. 

```{r}
nas <- cde_schls %>% 
  summarise(CDSCode = mean(is.na(CDSCode)), Short_CDS = mean(is.na(Short_CDS)),
            School = mean(is.na(School)), Latitude = mean(is.na(Latitude)), 
            Longitude = mean(is.na(Longitude)))

kable(nas)
```
  
## Write the File as a CSV 

That's it! Finally, we'll write the cleaned file as a csv. 

```{r}
write_csv(cde_schls, "data/cleaned_cde_schools.csv")
```
  
## Sample Plot: Top Ten Biggest School Districts 

```{r}
# Top ten biggest school districts in CA 

top_ten <- cde_schls %>% 
  count(District) %>% 
  filter(min_rank(-n) <= 10) %>% 
  arrange(desc(n))

ggplot(data = top_ten, aes(x = reorder(District, n), y = n)) +
  geom_bar(stat = "identity", fill = "cyan4", alpha = .75) + 
  coord_flip() + 
  labs(title = "Top Ten Biggest School Districts in California", 
       subtitle = "Data: California Department of Education",
       x = "",
       y = "Schools")
```

## Sample Plot: Interactive Map of Ten Oldest Schools 

```{r message = F, warning = F}
# A map of the ten oldest schools in CA  

library(leaflet)
library(lubridate)

ten_early <- cde_schls %>% 
  select(School, Date = OpenDate, lat = Latitude, lon = Longitude) %>% 
  mutate(label = paste0(year(Date), ": ", School)) %>% 
  filter(min_rank(Date) <= 10) %>% 
  arrange(Date)

ten_early %>% 
  leaflet() %>% 
  addTiles() %>% 
  addMarkers(popup = ten_early$label)
```

