---
title: "How to Use the CDE Public Schools Directory"
author: "Ryan Estrellado"
date: "5/20/2017"
output: ioslides_presentation
---

## The Problem 

The California Department of Education makes a dataset of public schools 
available [on their website](http://www.cde.ca.gov/ds/si/ds/pubschls.asp), but the dataset can be a little hard to use. In 
particular, CDE uses a fourteen digit school ID. This makes it inconvenient to 
merge with datasets like CASEMIS and SEIS, which use the shorter seven digit 
school ID. 

## The Problem (continued)  

The directory of active schools contains over 11,000 schools, so doing this 
transformation by hand can be time and processor intensive using more 
conventional tools. 

## Proposed Solution 

Using a script to clean and output the dataset accomplishes two things: 

- It efficiently and accurately transforms the fourteen digit school ID code to 
the more commonly used seven digit ID code 
- Since the script reads the data in directly from the CDE website, the script 
makes it very easy to download and clean a fresh version whenever needed  
- The cleaned dataset can be hosted on GitHub for others to use 

## Use Cases 

Once the dataset is cleaned, it can be used in a variety of ways. Some examples 
are: 

- Using it as a lookup table in Excel or Access to efficiently convert school 
IDs in datasets like CASEMIS into actual school names  
- Building interactive maps to evaluate and plan itinerant service provider caseloads, classrooms, or any system that requires the visualization of school 
locations 

## The Original Dataset  

The original dataset has 49 columns. Note the fourteen digit code in the 
`CDSCode` field. 

```{r message = F, echo = F, warning = F, comment = NULL}
# Load the required packages

library(tidyverse)
library(knitr)
library(stringr)
```

```{r echo = F}
opts_chunk$set(message = F, echo = F, warning = F, comment = NULL)
```

```{r}
original <- read_tsv("ftp://ftp.cde.ca.gov/demo/schlname/pubschls.txt")  

head(original, n = 5)
```

## The Cleaned Dataset  

The cleaned dataset removes 21 columns and adds a column for the seven digit 
school ID, `Short_CDS`: 

```{r}
cde_schls <- original %>% 
  filter(StatusType == "Active") %>% 
  select(-c(AdmFName1:AdmEmail3, FundingType, Magnet, StatusType, Street:State, DOC, CharterNum, SOC, EdOpsCode, EILCode))

cde_schls <- cde_schls %>% 
  mutate(Short_CDS = str_sub(CDSCode, 8, 20)) %>% 
  select(CDSCode, Short_CDS, everything()) 

head(cde_schls, n = 5)
```

## Download the Cleaned Dataset 

You can download the cleaned dataset by right clicking and selecting download 
linked file: 

[cleaned_cde_schools.csv](https://github.com/restrellado/california-public-schools/raw/master/data/cleaned_cde_schools.csv) 

## Ways to Use the Data 

The following are some examples of how this data can be used. 

## Top Ten Biggest Districts 

```{r}
# Top ten biggest school districts in CA 

top_ten <- cde_schls %>% 
  count(District) %>% 
  filter(min_rank(-n) <= 10) %>% 
  arrange(desc(n))

ggplot(data = top_ten, aes(x = reorder(District, n), y = n)) +
  geom_bar(stat = "identity", fill = "cyan4", alpha = .75) + 
  coord_flip() + 
  labs(title = "Top Ten Biggest School Districts in California", 
       subtitle = "Data: California Department of Education",
       x = "",
       y = "Schools")
```

## Top Ten Oldest Schools 

Click on the markers for more information. 

```{r}
# A map of the ten oldest schools in CA  

library(leaflet)
library(lubridate)

ten_early <- cde_schls %>% 
  select(School, Date = OpenDate, lat = Latitude, lon = Longitude) %>% 
  mutate(label = paste0(year(Date), ": ", School)) %>% 
  filter(min_rank(Date) <= 10) %>% 
  arrange(Date)

ten_early %>% 
  leaflet() %>% 
  addTiles() %>% 
  addMarkers(popup = ten_early$label)
```

## Service Provider Caseloads

This map uses simulated data. The size of the circle represents the size of the
caseload. 

```{r}
library(leaflet)
set.seed(2)

pick <- sample(c(1:length(cde_schls)), 10, replace = F)

schools <- cde_schls$School[pick]
lat <- cde_schls$Latitude[pick]
lon <- cde_schls$Longitude[pick]
caseload <- sample(c(15:50), 10, replace = T)
providers <- paste0("Some DIS Provider ", c(1:10), " ", "caseload: ", caseload) 

clp <- data.frame(schools, caseload, providers, lat, lon) 

clp %>% 
  leaflet() %>% 
  addTiles() %>% 
  addCircleMarkers(popup = clp$providers, radius = .50 * (clp$caseload))
```

